---
title: "Lecture – Working with uncertainty in dairy science"
subtitle: "ANSCI 4040 – Dairy Data Analysis & Analytics"
author: "Ass. Prof. Dr. Miel Hostens"
footer: "https://bovi-analytics.github.io/DigitalDairyManagementandDataAnalytics/"
logo: "images/Bovi-Analytics-SmallLogo.png"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

## Big idea

**Every number you see on a farm is an estimate.**

-   It has **sampling uncertainty** (finite cows, finite days)
-   It has **measurement uncertainty** (sensor error, missingness)
-   It has **model uncertainty** (assumptions, misspecification)

Today we learn how to **reason, quantify, and communicate** that uncertainty.

## Learning goals (part 1)

1.  Translate a dairy question into **(data type → model → test)**

2.  Run and interpret:

    -   a hypothesis test for a **normal** outcome (mean difference)
    -   a hypothesis test for a **binomial** outcome (proportion difference)

3.  Explain the **true value** of a *p*-value (what it is / is not)

## Learning goals (part 2)

4.  Distinguish **statistically different** vs **biologically different**

5.  Distinguish **association** vs **prediction**, and interpret **Se/Sp/PPV/NPV/Accuracy**

::: notes
Key teaching emphasis: Students often confuse “significant” with “important” and “correlation” with “usable prediction”.
:::

# Part 1 — Uncertainty: where it enters the dairy pipeline

## Where does uncertainty come from?

**A. Sampling** (finite cows, finite events)

Example: pregnancy rate on 25 AIs is noisier than on 250 AIs

**B. Measurement** (sensors, humans, lab assays)

Example: activity sensor misses mild estrus; SCC lab has CV

## Where does uncertainty come from?

**C. Model** (assumptions)

Example: t-test assumes approximate normality / equal variances

**D. Decision context** (costs and consequences)

Example: a mastitis alert that fires too often wastes time

::: notes
Bring this back later: statistical test choice is about *data-generating process*.
:::

## “Uncertainty” is not the enemy

In dairy science we often use data to:

-   **Understand** (association, mechanism)
-   **Decide** (intervention, management)
-   **Predict** (risk scores, alerts)

Uncertainty can be:

-   **Quantified** (SE, CI, posterior)
-   **Reduced** (better design, more cows, better sensors)
-   **Managed** (thresholds, decision analysis)

# Part 2 — P-values: what they are, and what they are NOT

## What a p-value *is*

A *p*-value is:

> The probability (under the **null hypothesis**) of observing a result **as extreme or more extreme** than what you observed.

So it is a statement about **data under a model**, not about “truth”.

## What a p-value is *not*

A *p*-value is **NOT**:

-   The probability that the null hypothesis is true
-   The probability your result will replicate
-   A measure of effect size or biological relevance
-   A “proof” of causality

::: notes
Students often interpret p=0.03 as “97% chance the treatment works”. That is wrong.
:::

## Why reporting *exact* p-values matters

When reporting results:

-   Prefer **exact p-values** (e.g., *P* = 0.003; *P* = 0.71) rather than only “*P* \< 0.05”.
-   Prefer **effect sizes + confidence intervals** around effects.

This matches good reporting guidance in dairy research.

*Example reporting guidance:* quantify p-values to \<0.001 and prefer 95% confidence intervals around effect estimates. (Lean et al., 2016)\[doi:10.3168/jds.2015-9445\]

::: notes
This point is explicitly recommended in the attached JDS invited review.
:::

# Part 3 — From question → data type → model → test

## Decision tree: which test?

```{mermaid}
flowchart TD
    Q["Start: What is your outcome?"] --> A{"Outcome type"}

    A -->|Continuous| C{"Independent groups"}
    A -->|Binary 0-1| B{"Groups"}
    A -->|Counts| K{"Rate or exposure"}
    A -->|Time-to-event| S["Survival model: Kaplan-Meier or Cox"]

    C -->|2 independent groups| C2{"Approx normal?"}
    C -->|Paired or repeated| CP["Paired t-test or mixed model"]
    C -->|3 or more groups| CA["ANOVA or linear model"]

    C2 -->|Yes| CT["t-test (Welch if variances differ)"]
    C2 -->|No| CW["Wilcoxon, transform, or robust"]

    B -->|1 proportion vs target| B1["Binomial test"]
    B -->|2 proportions| B2{"Expected counts large?"}
    B2 -->|Yes| B2a["Chi-square or two-proportion z test"]
    B2 -->|No| B2b["Fisher exact test"]
    B -->|Many covariates| BL["Logistic regression or mixed logistic"]

    K -->|Counts per time/unit| KP{"Overdispersed?"}
    KP -->|No| K1["Poisson regression"]
    KP -->|Yes| K2["Negative binomial regression"]
    K -->|Many zeros| KZ["Zero-inflated or hurdle models"]

    S --> S2{"Clustering (herd or cow)?"}
    S2 -->|Yes| S3["Frailty or mixed-effects survival"]
    S2 -->|No| S4["Standard survival analysis"]
```

::: notes
This slide is your “visual”. Emphasize: pick the test by (1) outcome distribution, (2) design/independence, (3) sample size and assumptions.
:::

## Data science × statistics: the handshake

Statistics helps you:

-   Choose a model aligned with the data-generating process

-   Quantify uncertainty (SE/CI)

-   Avoid false discoveries / overconfidence

## Data science × statistics: the handshake

Data science helps you:

-   Build reproducible pipelines

-   Validate models out-of-sample

-   Scale to large, messy farm data

**Best practice:** combine both.

# Part 4 — Hypothesis testing examples

## Example 1 (Normal): milk yield change vs a target

**Question:** A farm says the new ration should increase 30-d milk yield by **+4 lbs/day**. We sample 24 cows and compute their *change* (after–before).

-   Outcome: continuous (lbs/day)
-   Design: paired (same cows)
-   Model: approx normal differences
-   Test: **paired t-test**

## Code (R)

```{r}
#| echo: true
n <- 24
true_effect <- 2.4     # true mean change (lbs/d)
sd_change <- 3.0

change <- rnorm(n, mean=true_effect, sd=sd_change)

# H0: mean change = 4 lbs/d (the claim) HA: mean change != 4
res <- t.test(change, mu = 4)
res
```

::: notes
This is a deliberately realistic scenario: lots of biological noise. Ask: even if p\>0.05, what is the CI around the effect?
:::

## Interpreting the result: emphasize effect size

The test checks whether the mean of the variable change is equal to 4.

Hypotheses:

-   H0: The mean = 4

-   H1: The mean ≠ 4

## What does the t-value mean?

**t = -3.7808**

-   The negative sign means the sample mean is **below 4**
-   The larger the absolute value of t, the stronger the evidence against H0

This suggests the mean is far from 4.

## Degrees of Freedom

**df = 23**

This means:

-   Number of observations = 24\
    (because df = n − 1)

## The *P-value*

***P*****-value = 0.0009681**

What this means:

-   The probability of getting this result if the true mean were 4 is very small

-   Since 0.0009681\< 0.05, the result is **statistically significant**

Conclusion:

We reject the null hypothesis.

## Confidence Interval

**95% Confidence Interval:**

1.077941 → 3.144723

Important:

-   The value **4 is not inside the interval**

-   This confirms that the mean is significantly different from 4

## Sample Mean

**Mean of change = 2.111332**

Interpretation:

-   The average value is about **2.11**

-   This is clearly lower than 4

## Final Conclusion

The data show that:

-   The mean of *change* is significantly different from 4

-   The true mean is most likely between **1.08 and 3.14**

-   The sample mean is **2.11**

**Final conclusion:**

-   The average change is significantly lower than 4.

-   Is it **biologically meaningful** for the farm?

------------------------------------------------------------------------

## Example 2 (Binomial): conception rate vs a baseline

**Question:** In a timed-AI protocol, baseline **P/AI = 0.35**. After a management tweak we observe **x = 18 pregnancies out of n = 40 AIs**.

-   Outcome: pregnancy (0/1)
-   Model: binomial
-   Test: **binomial test** (one-sample proportion)

## Example 2 (Binomial): conception rate vs a baseline

```{r}
#| echo: true
x <- 18
n <- 40
p0 <- 0.35

#Binomial test
binom.test(x, n, p = p0, alternative = "two.sided")
```

::: notes
Highlight: with n=40, uncertainty is still big. Ask students to inspect CI for the proportion.
:::

## Example 3 (Two groups): pregnancy probability in 2 treatments

**Question:** Treatment A vs B; outcome is pregnant yes/no.

-   If expected counts are large → chi-square / two-proportion test
-   If expected counts are small → Fisher exact

## Example 3 (Two groups): pregnancy probability in 2 treatments

```{r}
#| echo: true
# Example counts
# A: 55 preg / 145 not
# B: 70 preg / 130 not

tab <- matrix(c(55,145,70,130), nrow=2, byrow=TRUE)
colnames(tab) <- c("Preg", "Not")
rownames(tab) <- c("A", "B")
tab
```

## Example 3 (Two groups): pregnancy probability in 2 treatments

```{r}
#| echo: true
#Chi-square
prop.test(tab)
```

## Example 3 (Two groups): pregnancy probability in 2 treatments

```{r}
#| echo: true
#Fisher exact
fisher.test(tab)
```

::: notes
Teach students: prop.test uses approximate methods; fisher is exact for small counts.
:::

# Part 5 — Beyond normal & binomial: common distributions in dairy

## Quick map: outcome → typical distribution

-   **Milk yield (daily)**: roughly normal after removing extremes; often needs mixed models
-   **SCC**: often log-normal (or gamma) after log transform
-   **Clinical mastitis events per lactation**: Poisson-like, but often overdispersed → negative binomial
-   **Time to pregnancy**: survival/time-to-event
-   **Proportions** (e.g., % infected): beta distribution on (0,1)

::: notes
You can show raw SCC vs log(SCC) histograms to motivate transformations.
:::

## Example: SCC looks log-normal (why we log-transform)

```{r}
set.seed(10)

# Simulate SCC-like data (cells/mL) with a log-normal shape
n <- 800
scc <- rlnorm(n, meanlog = log(120000), sdlog = 0.8)

par(mfrow=c(1,2))
hist(scc, breaks=30, main="Raw SCC", xlab="cells/mL")
hist(log10(scc), breaks=30, main="log10(SCC)", xlab="log10 cells/mL")
par(mfrow=c(1,1))
```

::: notes
Connect this to model assumptions: many tests/regressions behave better after transformation.
:::

## Visual: same mean, different uncertainty

```{r}
set.seed(1)

# Compare sampling uncertainty for a proportion with different n
p <- 0.35
n_small <- 20
n_large <- 200

x_small <- rbinom(5000, n_small, p)/n_small
x_large <- rbinom(5000, n_large, p)/n_large

par(mfrow=c(1,2))
hist(x_small, breaks=20, main="Estimated P/AI, n=20", xlab="p-hat")
hist(x_large, breaks=20, main="Estimated P/AI, n=200", xlab="p-hat")
par(mfrow=c(1,1))
```

## Poisson vs negative binomial

**Question:** Count of clinical mastitis cases per cow-year.

-   Poisson assumes mean = variance
-   Dairy health data often have variance \> mean (**overdispersion**) → negative binomial

## Poisson vs negative binomial

```{r}
set.seed(2)

# Simulate overdispersed counts
n <- 300
mu <- 0.6
# Negative binomial with size parameter (smaller = more dispersion)
size <- 0.8
counts <- rnbinom(n, mu=mu, size=size)

mean(counts)
var(counts)

hist(counts, breaks=0:max(counts), main="Mastitis cases per cow-year", xlab="count")
```

## Time-to-pregnancy is survival data

```{r}
set.seed(12)
library(survival)

N <- 250

# --- Define cure fraction (40% never pregnant) ---
cure_prob <- 0.40
is_cured <- rbinom(N, 1, cure_prob)  # 1 = will never conceive

# --- Generate event times for the susceptible group ---
# Use Weibull to control the shape (delay + decreasing hazard)
# Shape >1 gives increasing hazard, scale controls timing
shape <- 3       # sharper increase after ~60 days
scale <- 80      # controls where the drop starts
event_time <- rweibull(N, shape = shape, scale = scale) + 60  # delay ~60 days

# Assign large times to cured individuals
true_time <- ifelse(is_cured == 1, 999, event_time)

# Administrative censoring at 200 days
censor <- 200
obs_time <- pmin(true_time, censor)
status <- ifelse(true_time <= censor, 1, 0)

# Kaplan–Meier curve
fit <- survfit(Surv(obs_time, status) ~ 1)

plot(fit,
     xlab = "Days",
     ylab = "Open cows",
     main = "Time to pregnancy",
     xlim = c(0, 200))
grid()
```

::: notes
Use this slide to reinforce: for time-to-event outcomes, survival analysis is the right tool and avoids bias from excluding cows that never conceive.
:::

# Part 6 — Statistical vs biological difference

## Why “statistically significant” can be misleading

You can get *P* \< 0.05 when:

-   The effect is tiny but **n is huge**
-   The effect is real but **irrelevant** for welfare, economics, or practicality

You can also get *P* \> 0.05 when:

-   The effect might be meaningful but **n is small**
-   There is high biological variability

## Why “statistically significant” can be misleading

So always report:

-   **Effect size**
-   **Confidence interval**
-   **Context-specific threshold** (“what difference matters?”)

## Practical tool: define a “biological minimum important difference”

Examples (illustrative):

-   Milk yield: +2 lbs/d might matter economically on some farms

-   Conception rate: +3 to +5 percentage points might matter if sustained

-   Disease incidence: -2% incidence might matter if treatment cost is high

::: notes
Let students propose what “biologically meaningful” means for one KPI.
:::

## Sample size: how many cows do we need?

Effect size you care about drives sample size.

From the invited review on reproductive intervention studies:

-   To detect an increase in pregnancy probability per insemination from **35% to 40%** with **power = 0.8** and **α = 0.05**, requires **1,471 cows per group**.
-   Additional cows may be required when there is **clustering** (e.g., cows within herds). (Lean et al., 2016)\[doi:10.3168/jds.2015-9445\]

# Part 7 — Association vs prediction

## Association answers “is X related to Y?”

-   Goal: inference, explanation, mechanism
-   Output: coefficients, CI, p-values
-   Example: higher rumination is associated with lower disease odds

## Prediction answers “can we forecast Y for a new cow?”

-   Goal: accurate out-of-sample performance
-   Output: sensitivity, specificity, PPV/NPV, ROC-AUC, calibration
-   Example: build an early-lactation alert for ketosis

**A model can have association without useful prediction.**

::: notes
Key message: prediction needs validation on new data and pays attention to prevalence and decision thresholds.
:::

## A quick demonstration: strong association, weak prediction

If an outcome is rare, even a decent model can have low PPV.

-   Prevalence: 5% disease
-   Sensitivity: 80%
-   Specificity: 90%

## A quick demonstration: strong association, weak prediction

Let’s compute PPV/NPV.

```{r}
#| echo: true
prev <- 0.05
se <- 0.80
sp <- 0.90

ppv <- (se*prev) / (se*prev + (1-sp)*(1-prev))
npv <- (sp*(1-prev)) / ((1-se)*prev + sp*(1-prev))

ppv; npv
```

## Confusion matrix (definitions)

Let:

-   TP = true positives
-   FP = false positives
-   TN = true negatives
-   FN = false negatives

Then:

-   **Sensitivity (Se)** = TP / (TP + FN)
-   **Specificity (Sp)** = TN / (TN + FP)
-   **Accuracy** = (TP + TN) / total
-   **PPV** = TP / (TP + FP)
-   **NPV** = TN / (TN + FN)

**PPV/NPV depend strongly on prevalence.**

## Worked dairy example: mastitis alert

Assume 1,000 cows monitored in a period:

-   True mastitis prevalence = 8% → 80 true cases
-   Algorithm: Se = 75%, Sp = 92%

Compute counts:

```{r}
N <- 1000
prev <- 0.08
se <- 0.75
sp <- 0.92

D <- round(N*prev)
ND <- N - D

TP <- round(se*D)
FN <- D - TP
TN <- round(sp*ND)
FP <- ND - TN

c(TP=TP, FP=FP, TN=TN, FN=FN)

accuracy <- (TP+TN)/N
ppv <- TP/(TP+FP)
npv <- TN/(TN+FN)

c(accuracy=accuracy, PPV=ppv, NPV=npv)
```

-   If the algorithm flags a cow as having mastitis, there’s about a 45% chance it actually has mastitis.

-   So negatives are much more reliable than positives here.

::: notes
Discuss: Is PPV high enough to justify an alert? What is the cost of FP vs FN on this farm?
:::

------------------------------------------------------------------------

# Part 8 — A light touch: regression and ANOVA

## Why regression shows up everywhere

A linear regression is:

-   A model for **continuous** outcomes
-   With predictors (continuous or categorical)

ANOVA is a special case:

-   Same linear model
-   Predictors can be categorical group indicators

## Lactation yield vs DIM and parity

```{r}
set.seed(3)

n <- 200
DIM <- sample(1:305, n, replace=TRUE)
parity <- factor(sample(c(1,2,3), n, replace=TRUE))

# Simulate a simple relationship
milk <- 35 + 0.02*(150 - abs(DIM-150)) +
  ifelse(parity=="1", -2, ifelse(parity=="2", 0, 1)) +
  rnorm(n, 0, 3)

d <- data.frame(milk, DIM, parity)

fit <- lm(milk ~ DIM + parity, data=d)
summary(fit)
```

# References

-   Lean, I.J. et al. (2016). *Recommendations for reporting intervention studies on reproductive performance in dairy cattle: Improving design, analysis, and interpretation of research on reproduction.* Journal of Dairy Science 99:1–17. doi:10.3168/jds.2015-9445.
